A Process for Emulating Comparative Oncology Trials with Real-world Evidence Studies (ENCORE): Process Development and Methodological Considerations for Oncology Real-World Data

### 2025-03-30 manuscript/manuscript.tex
added updated word count

\ul{\textbf{Competing Interests Statement:}} Dr.~Weberpals is now an
employee of AstraZeneca and owns stocks in AstraZeneca. Dr.~Kehl has
received research funding from Meta, Inc.~to his institution. Drs.
Espirito and Robert are employees of McKesson and own McKesson stock.
Dr.~Wang has consulted ad hoc for Exponent Inc.~and MITRE a federally
funded research center for the Centers for Medicare and Medicaid
Services on unrelated work. Dr.~Schneeweiss is participating in
investigator-initiated grants to the Brigham and Women's Hospital from
Bayer and UCB unrelated to the topic of this study. He consults for and
owns equity in Aetion Inc., a software manufacturer. He is an advisor to
Temedica GmbH, a patient-oriented data generation company. His interests
were declared, reviewed, and approved by the Brigham and Women's
Hospital in accordance with their institutional compliance policies.

@@

\emph{Manuscript last updated: 2025-03-30 15:14:45.088413}

@@

Randomized controlled trials (RCTs) are the go-to methodology for
establishing the efficacy and safety of medical products. Under the
21\textsuperscript{st} Century Cures Act,\textsuperscript{1} the Food
and Drug Administration (FDA) established a framework to factor
consideration of real-world evidence (RWE) generated from routine-care
health data such as electronic health records (EHR) to evaluate and
contextualize the safety and effectiveness of novel cancer
therapies.\textsuperscript{2} Accounting for 30\% of new drug approvals,
oncology was the disease area with the most FDA drug approvals in
2024\textsuperscript{3} and has many areas of high unmet medical need to
treat serious conditions. Therefore, RWE has particularly important
potential to complement evidence coming from RCTs in the field of
oncology. Potential use cases include the assessment of effectiveness in
specific patient populations that are underrepresented in RCTs, the
construction of external control arms in single-arm trials that have
feasibility or equipoise challenges, or the precision oncology-focused
discovery of biomarkers among pan-tumor populations that harbor specific
genomic and immuno-pathological signatures.

However, the validity and transportability of results derived between
RWE studies and RCTs depends on multiple factors. Frequently referenced
limitations include, lack of baseline randomization, missing data, small
study sizes, data discontinuity,\textsuperscript{4, 5} adoption of
changes in guidelines in real-world care and the inability to measure
and emulate common eligibility criteria, including prognostic factors,
and standardized response assessments in real-world data
(RWD).\textsuperscript{6} While examples of oncology trial emulations
have been published,\textsuperscript{6--8} a systematic and scaled
approach to emulate a diverse set of different oncology trials with
multiple heterogeneous databases is necessary to gain confidence in the
validity of RWE studies, evaluate regulatory considerations, and to
provide context as to which questions can be validly answered.

The RCT DUPLICATE initiative\textsuperscript{9} increased our
understanding of when RWE studies can come to causal conclusions on
treatment effects by benchmarking results against RCTs under the
assumption that each RCT finding reflects a causal treatment effect. In
settings where the RCT designs could be emulated well, RWE studies were
able to reach the same conclusions.\textsuperscript{10} However, this
prior work from RCT-DUPLICATE focused primarily on emulating trials in
non-oncology settings using claims databases.

@@

The rationale and operationalization of each criterion is listed in
detail in Table~\ref{tbl-criteria}. We will mainly consider pivotal
large late-phase interventional RCTs after 2011 because treatment
guidelines among included cancer indications have undergone significant
changes in recent years. Due to the rapid adoption of breakthrough
therapies in routine care, it is unlikely to find patients who may be
still treated with outdated treatment regimens in current clinical
practice. Conversely, we excluded trials with results published too
recently. To allow for enough data and follow-up time accrual in the
real-world databases used for this project. Although there have been
substantial methodological advancements to increase our understanding on
the emulation and comparison of real-world progression-free survival
(PFS) and objective response rates (ORR) to a
RECISTv1.1\textsuperscript{15}-based PFS and ORR assessment in
RCTs\textsuperscript{16, 17}, imaging-based evaluations still hold a
level of granularity which may not is often not reflected in
chart-abstracted assessments of a patient's progression in routine
care.\textsuperscript{18, 19} The timing and cadence of intervals
between progression assessments can differ between RCTs and routine care
which may result in measurement error and bias.\textsuperscript{20}
Given this, and the large number of other methodological challenges
including missing data, small sample sizes, data discontinuity and
rapidly changing guideline treatments, we focus on the emulation of
overall survival (OS) as the endpoint of interest. Therefore, we include
only trials that have reported overall survival (OS) as one of the
pre-specified endpoints in the protocol.

>>>>

\ul{\textbf{Competing Interests Statement:}} Dr.~Weberpals is now an
employee of AstraZeneca and owns stocks in AstraZeneca. Dr.~Kehl has
received research funding from Meta, Inc.~to his institution. Drs.
Espirito and Robert are employees of McKesson and own McKesson stock.
Dr.~Wang has consulted ad hoc for Exponent Inc.~and MITRE a federally
funded research center for the Centers for Medicare and Medicaid
Services on unrelated work. Dr.~Glynn has received support for
investigator-initiated grants to the Brigham and Women's Hospital from
Amarin, AstraZeneca, Kowa, Novartis, and Pfizer unrelated to the current
work. Dr.~Schneeweiss is participating in investigator-initiated grants
to the Brigham and Women's Hospital from Bayer and UCB unrelated to the
topic of this study. He consults for and owns equity in Aetion Inc., a
software manufacturer. He is an advisor to Temedica GmbH, a
patient-oriented data generation company. His interests were declared,
reviewed, and approved by the Brigham and Women's Hospital in accordance
with their institutional compliance policies.

@@

\emph{Manuscript last updated: 2025-03-30 19:11:35.894604}

@@

Randomized controlled trials (RCTs) are the go-to methodology for
establishing the efficacy and safety of medical products. Directed by
the 21\textsuperscript{st} Century Cures Act,\textsuperscript{1} the
Food and Drug Administration (FDA) established a framework to factor
consideration of real-world evidence (RWE) generated from routine-care
health data such as electronic health records (EHR) to evaluate and
contextualize the safety and effectiveness of novel cancer
therapies.\textsuperscript{2} Accounting for 30\% of new drug approvals,
oncology was the disease area with the most FDA drug approvals in
2024\textsuperscript{3} and has many areas of high unmet medical need to
treat serious conditions. Therefore, RWE has particularly important
potential to complement evidence from RCTs in the field of oncology.
Potential use cases include the assessment of effectiveness in specific
patient populations that are underrepresented in RCTs, the construction
of external control arms in single-arm trials that have feasibility or
equipoise challenges, or the precision oncology-focused discovery of
biomarkers among pan-tumor populations that harbor specific genomic and
immuno-pathological signatures.

However, the validity and transportability of results derived between
RWE studies and RCTs depends on multiple factors. Frequently referenced
limitations include lack of baseline randomization, missing data,
unmeasured confounding, small study sizes, data
discontinuity,\textsuperscript{4, 5} adoption of changes in guidelines
in real-world care and the inability to measure and emulate common
eligibility criteria, including prognostic factors, and standardized
response assessments in real-world data (RWD).\textsuperscript{6} While
examples of oncology trial emulations have been
published,\textsuperscript{6--8} a systematic and scaled approach to
emulate a diverse set of oncology trials with multiple heterogeneous
databases is necessary to gain confidence in the validity of RWE
studies, evaluate regulatory considerations, and to provide context as
to which questions can be validly answered.

The RCT DUPLICATE initiative\textsuperscript{9} increased our
understanding of when RWE studies can come to causal conclusions on
treatment effects by benchmarking results against RCTs under the
assumption that each RCT finding reflects a causal treatment effect. In
settings where the RCT designs could be emulated well, RWE studies were
able to reach similar conclusions.\textsuperscript{10} However, this
prior work from RCT-DUPLICATE focused primarily on emulating trials in
non-oncology settings using claims databases.

@@

The rationale and operationalization for each criterion is listed in
detail in Table~\ref{tbl-criteria}. We will mainly consider pivotal
large late-phase interventional RCTs after 2011 because treatment
guidelines among included cancer indications have undergone significant
changes in recent years. Due to the rapid adoption of breakthrough
therapies in routine care, few patients currently initiate outdated
treatment regimens in current clinical practice. Conversely, we excluded
trials with results published too recently to allow for enough data and
follow-up time accrual in the real-world databases used for this
project. We did not define a global cut-off as the requirements for
follow-up time are different for each cancer type and population (e.g.,
advanced NSCLC versus early-stage breast cancer) and the decisions were
made on a case-by-case basis.

Although there have been substantial methodological advancements to
increase our understanding on the emulation and comparison of real-world
progression-free survival (PFS) and objective response rates (ORR) to a
RECISTv1.1\textsuperscript{15}-based PFS and ORR assessment in
RCTs\textsuperscript{16, 17}, imaging-based evaluations still hold a
level of granularity which is not well reflected in chart-abstracted
assessments of a patient's progression in routine
care.\textsuperscript{18, 19} The timing and cadence of intervals
between progression assessments can differ between RCTs and routine care
which may result in measurement error and bias.\textsuperscript{20}
Given this, and the large number of other methodological challenges
including missing data, small sample sizes, data discontinuity and
rapidly changing guideline treatments, we focus on the emulation of
overall survival (OS) as the endpoint of interest. Therefore, we include
only trials that have reported overall survival (OS) as one of the
pre-specified endpoints in the protocol.
